QuickSched: Task-based parallelism with dependencies and conflicts

### 2015-10-20 paper/paper.tex
first round of corrections, sections 1--3. will update validation and results next.

\begin{abstract}
This paper describes QuickSched, a compact and efficient Open-Source
C-language library for task-based shared-memory parallel programming.
QuickSched extends the standard dependency-only scheme of task-based
programming with the concept of task conflicts, i.e.~sets of tasks
that can be executed in any order, yet not concurrently.
These conflicts are modelled using exclusively lockable
hierarchical resources.
The scheduler is shown to perform and scale well on a 64-core parallel
shared-memory machine for two example problems: A tiled QR
decomposition and a task-based Barnes-Hut tree code.
\end{abstract}

@@

\fig{Tasks} shows such a DAG for a set of tasks.
The arrows indicate the direction of the dependency, i.e.~an
arrow from task $A$ to task $B$ indicates that task $B$ depends
on task $A$.
In a parallel setting, tasks $A$, $G$, and $J$ can be
executed concurrently.
Once task $G$ has completed, tasks $F$ and $H$ become available
and can be executed by any other computational thread.

@@

This paper presents QuickSched, a framework for task-based
parallel programming with constraints, which aims to achieve
the following goals:
\begin{itemize}
    \item {\em Correctness}: All constraints, i.e.~dependencies and
        conflicts, must be correctly enforced,
    \item {\em Speed}: The overheads associated with task management
        should be as small as possible,
    \item {\em Memory/cache efficiency}: Tasks accessing similar
        sets of data should be preferentially executed on the
        same core to preserve memory/cache locality as far as possible, and
    \item {\em Parallel efficiency}: Tasks should be executed in an order
        that sufficient work is available for all computational
        threads at all times.
\end{itemize}
\noindent 
Section~2 describes the main design considerations, Section~3 the
underlying algorithms and data structures, and
Section~4 their specific implementation in QuickSched.
Section~5 presents two test-cases:
\begin{enumerate}
    \item The tiled QR
    decomposition described in \cite{ref:Buttari2009} and for
    which the QUARK scheduler was originally developed, and
    \item A task-based Barnes-Hut tree-code to compute the
    gravitational N-body problem similar to the FMM codes
    of \cite{ref:Ltaief2012} and \cite{ref:Agullo2013},
\end{enumerate}
These real-world examples show how QuickSched can be used in practice,
and can be used to assess its efficiency.
Section~6 concludes with some general observations and future work
directions.

@@

From a programmer's perspective, there are two main paradigms for generating
task dependencies:
\begin{itemize}
  \item Implicitly via spawning and waiting, e.g.~as is done in Cilk
    \citep{ref:Blumofe1995}, or
  \item Automatic extraction from data dependencies, e.g.~as is done in OmpSs
    \citep{ref:Duran2011}.
\end{itemize}

@@

In the examples presented herein, we have chosen our task decomposition
and granularity such that
\begin{itemize}
  \item Each task maximizes the ratio of computation to data required,
  \item The resources required for each task fit comfortably in the
    lowest-level caches of the underlying system.
\end{itemize}
\noindent The first critera is biased towards bigger tasks, while the
second limits their size.
The parameters controlling the size of the tasks in the examples, 
i.e.~the tile size in the QR decomposition and the limits $n_\mathsf{max}$
and $n_\mathsf{task}$ were determined empirically and only optimized
to the closest power of two or rough power of ten, respectively.

@@

What the task does is determined by the {\tt type}
field, e.g.~which can be mapped to any particular function,
and the {\tt data} pointer which points to an array of
{\tt size\_data} bytes containing data specific to the task,
e.g.~the parameters for a specific function call.
Both fields are application-specific and therefore not
important for the scheduler itself.

@@

Finally, {\tt cost} and {\tt weight} are measures
for the relative computational cost of this task, and the
relative cost of the critical path following the
dependencies of this task, respectively, i.e.~the task's cost
plus the maximum dependent task's weight (see \fig{TaskWeight}).
The task cost can be either a rough estimate provided by the user,
or the actual cost of the same task last time it was executed.
The task weights are computed by traversing
the tasks DAG in reverse topological order following their dependencies,
e.g.~as per \cite{ref:Kahn1962}, and computing each task's weight, e.g.
\begin{equation*}
  \mbox{weight}_i = \mbox{cost}_i + \max_{j \in \mbox{\small unlocks}_i}\left\{\mbox{weight}_j\right\}.
\end{equation*}
\noindent where $\mbox{weight}_i$ and $\mbox{cost}_i$ are the
weight and cost of the $i$th task, respectively, and 
$\mbox{unlocks}_i$ is the set of tasks that the $i$th task
unlocks.

@@

Unfortunately, there is no way of efficiently traversing all
the elements in the heap in decreasing order.
The array of tasks is therefore traversed as if it were sorted,
returning the first task that can be locked.
Although the first task in the array will be the task with
maximum weight, the following tasks are only loosely ordered,
where the $k$th of $n$ tasks has a larger weight than at least
$\lfloor n/k\rfloor -1$ other tasks.
Although this is not a particularly good lower bound, it turns
out to be quite sufficient in practice.

@@

Obtaining a task from the queue can be implemented as follows:
\begin{center}\begin{minipage}{0.9\textwidth}
    \begin{lstlisting}
struct task *queue_get(struct queue *q) {
  struct task *res = NULL;
  int j, k;
  while (atomic_cas(q->lock, 0, 1) != 0) {}
  for (k = 0; k < q->count; k++) {
    for (j = 0; j < q->tasks[k]->nr_locks; j++)
      if (!resource_lock(q->tasks[k]->lock[j])) break;
    if (j < q->tasks[k]->nr_locks)
      for (j = j - 1; j >= 0; j--)
        resource_unlock(q->tasks[k]->lock[j]);
    else
      break;
  }
  if (k < q->count) {
    res = q->tasks[k];
    q->tasks[k] = q->tasks[--q->count];
    trickle-down the kth entry of q->tasks.
  }
  q->lock = 0;
  return res;
}
    \end{lstlisting}
\end{minipage}\end{center}
\noindent where as with the queue insertion, the queue is first
locked for exclusive access (line~4).
The array of task pointers is then traversed (line~5),
locking the resources of each task (lines~6--7).
If any of these locks fail (line~8), the locks that were obtained
are released (lines~9--10), otherwise, the traversal is aborted
(line~12).
If all the locks on a task could be obtained (line~14), the
task pointer is replaced by the last pointer in the heap (line~16)
and the heap order is restored (line~17).
Finally, the queue lock is released (line~19) and the locked task
or, if no lockable task could be found, {\tt NULL} is returned.

@@

Note also that although protecting the entire queue with a mutex
is not particularly scalable, and several authors, e.g.~REFS,
have presented concurrent data structures that avoid this type
of locking.
However, since we normally use one queue per computational thread,
contention will only happens due to work-stealing, i.e.~when
another idle computational thread tries to poach tasks.
Since this happens only rarely, we opt for the simpler locking approach.

@@

The tasks are executed as follows:
\begin{center}\begin{minipage}{0.9\textwidth}
    \begin{lstlisting}
void qsched_run(qsched *s, void (*fun)(int, void *)) {
  qsched_start(s);
  #pragma omp parallel
  {
    int qid = omp_get_thread_num() % s->nr_queues;
    struct task *t;
    while ((t = qsched_gettask(s, qid)) != NULL) {
      fun(t->type, t->data);
      qsched_done(s, t);
    }
  }
}
    \end{lstlisting}
\end{minipage}\end{center}
\noindent where {\tt qsched\_start} initializes the tasks and
fills the queues (line~1).
For simplicity, OpenMP \citep{ref:Dagum1998}, which is available
for most compilers, is used to create a parallel section
in which the code between lines~4 and~11 is executed
concurrently.
A version using {\tt pthreads} \citep{ref:Pthreads1995}
directly\footnote{In most environments, OpenMP is implemented
on top of {\tt pthreads}, e.g. gcc's libgomp.} is also available.
The parallel section consists of a loop (lines~7--10) in
which a task is acquired via {\tt qsched\_gettask}
and its type and data are passed to a user-supplied
{\em execution function} {\tt fun}.
Once the task has been executed, it is returned to the
scheduler via the function {\tt qsched\_done}, i.e.~to
unlock its resources and unlock dependent tasks.
The loop terminates when the scheduler runs out of tasks,
i.e.~when {\tt qsched\_gettask} returns {\tt NULL}, and
the function exits once all the threads have exited their
loops.

>>>>

\begin{abstract}
This paper describes QuickSched, a compact and efficient Open-Source
C-language library for task-based shared-memory parallel programming.
QuickSched extends the standard dependency-only scheme of task-based
programming with the concept of task conflicts, i.e.~sets of tasks
that can be executed in any order, yet not concurrently.
These conflicts are modelled using exclusively lockable
hierarchical resources.
The scheduler itself prioritizes tasks along the critical path
of execution and is shown to perform and scale well on a 64-core parallel
shared-memory machine for two example problems: A tiled QR
decomposition and a task-based Barnes-Hut tree code.
\end{abstract}

@@

\fig{Tasks} shows such a DAG for a set of tasks with
arrows indicating the direction of the dependencies, i.e.~an
arrow from task $A$ to task $B$ indicates that task $B$ depends
on task $A$.
In a parallel setting, tasks $A$, $G$, and $J$ can be
executed concurrently.
Once task $G$ has completed, tasks $F$ and $H$ become available
and can be executed by any other computational thread.

@@

Several libraries provide some mechanism to model such
conflicts, either directly or indirectly.
In the QUARK scheduler, conflicts can be modeled by explicitly
marking dependencies as concurrent.
KAAPI and OmpSS, on the other hand, allow marking access to
certain variables as reductions, yet only for basic operations,
e.g.~summation or maximum/minimum.

This paper presents QuickSched, a framework for task-based
parallel programming with constraints, which aims to achieve
the following goals:
\begin{itemize}
    \item {\em Correctness}: All constraints, i.e.~dependencies and
        conflicts, must be correctly enforced,
    \item {\em Speed}: The overheads associated with task management
        should be as small as possible,
    \item {\em Memory/cache efficiency}: Tasks accessing similar
        sets of data should be preferentially executed on the
        same core to preserve memory/cache locality as far as possible, and
    \item {\em Parallel efficiency}: The order in which the tasks
        are executed should be chosen such
        that sufficient work is available for all computational
        threads at all times.
\end{itemize}
\noindent 
Section~2 describes the main design considerations, Section~3 the
underlying algorithms and data structures, and
Section~4 their specific implementation in QuickSched.
Section~5 presents two test-cases:
\begin{enumerate}
    \item The tiled QR
    decomposition described in \cite{ref:Buttari2009} and for
    which the QUARK scheduler was originally developed, and
    \item A task-based Barnes-Hut tree-code to compute the
    gravitational N-body problem similar to the FMM codes
    of \cite{ref:Ltaief2012} and \cite{ref:Agullo2013},
\end{enumerate}
These real-world examples show how QuickSched can be used in practice,
and can be used to assess its efficiency.
Section~6 concludes with some general observations and future work
directions.

@@

From a programmer's perspective, there are two main paradigms for generating
task dependencies:
\begin{itemize}
  \item Implicitly via spawning and waiting, e.g.~as is done in Cilk
    and OpenMP~3.0, or
  \item Automatic extraction from data dependencies, e.g.~as is done in
    StarPU, QUARK, and OmpSs.
\end{itemize}

@@

In the examples presented herein, we have chosen our task decomposition
and granularity such that
\begin{itemize}
  \item Each task maximizes the ratio of computation to data required,
  \item The resources required for each task fit comfortably in the
    lowest-level caches of the underlying system.
\end{itemize}
\noindent The first critera is biased towards bigger tasks, while the
second limits their size.
The parameters controlling the size of the tasks in the examples, 
i.e.~the tile size in the QR decomposition and the limits $n_\mathsf{max}$
and $n_\mathsf{task}$ were determined empirically and only optimized
to the closest power of two or rough power of ten, respectively.
Further tuning these parameters could very likely lead to further
performance gains, but such an effort would go beyond the scope,
and point, of this paper.

@@

{\em What} the task does is determined by the {\tt type}
field, e.g.~which can be mapped to any particular function,
and the {\tt data} pointer which points to an array of
{\tt size\_data} bytes containing data specific to the task,
e.g.~the parameters for a specific function call.
Both fields are application-specific and therefore not
important for the scheduler itself.

@@

Finally, {\tt cost} and {\tt weight} are measures
for the relative computational cost of this task, and the
relative cost of the critical path following the
dependencies of this task, respectively, i.e.~the task's cost
plus the maximum dependent task's weight (see \fig{TaskWeight}).
The task cost can be either a rough estimate provided by the user,
or the actual cost of the same task last time it was executed.
The task weights are computed by traversing
the tasks DAG in reverse topological order following their dependencies,
e.g.~as per \cite{ref:Kahn1962} in $\mathcal O(n)$ for $n$ tasks,
and computing each task's weight, e.g.
\begin{equation*}
  \mbox{weight}_i = \mbox{cost}_i + \max_{j \in \mbox{\small unlocks}_i}\left\{\mbox{weight}_j\right\}.
\end{equation*}
\noindent where $\mbox{weight}_i$ and $\mbox{cost}_i$ are the
weight and cost of the $i$th task, respectively, and 
$\mbox{unlocks}_i$ is the set of tasks that the $i$th task
unlocks.

@@

Unfortunately, there is no way of efficiently traversing all
the elements in such a heap in decreasing order.
The array of tasks is therefore traversed as if it were sorted,
returning the first task that can be locked.
Although the first task in the array will be the task with
maximum weight, the following tasks are only loosely ordered,
where the $k$th of $n$ tasks has a larger weight than at least
$\lfloor n/k\rfloor -1$ other tasks.
Although this is not a particularly good lower bound, it turns
out to be quite sufficient in practice.

@@

Obtaining a task from the queue can be implemented as follows:
\begin{center}\begin{minipage}{0.9\textwidth}
    \begin{lstlisting}
struct task *queue_get(struct queue *q) {
  struct task *res = NULL;
  int j, k;
  while (atomic_cas(q->lock, 0, 1) != 0) {}
  for (k = 0; k < q->count; k++) {
    for (j = 0; j < q->tasks[k]->nr_locks; j++)
      if (!resource_lock(q->tasks[k]->lock[j])) break;
    if (j < q->tasks[k]->nr_locks)
      for (j = j - 1; j >= 0; j--)
        resource_unlock(q->tasks[k]->lock[j]);
    else
      break;
  }
  if (k < q->count) {
    res = q->tasks[k];
    q->tasks[k] = q->tasks[--q->count];
    trickle-down the kth entry of q->tasks.
  }
  q->lock = 0;
  return res;
}
    \end{lstlisting}
\end{minipage}\end{center}
\noindent where, as with the queue insertion, the queue is first
locked for exclusive access (line~4).
The array of task pointers is then traversed (line~5),
locking the resources of each task (lines~6--7).
If any of these locks fail (line~8), the locks that were obtained
are released (lines~9--10), otherwise, the traversal is aborted
(line~12).
If all the locks on a task could be obtained (line~14), the
task pointer is replaced by the last pointer in the heap (line~16)
and the heap order is restored (line~17).
Finally, the queue lock is released (line~19) and the locked task,
or {\tt NULL} if no lockable task could be found, is returned.

@@

Note also that protecting the entire queue with a mutex
is not particularly scalable, and several authors, e.g.~\cite{ref:Sundell2003},
have presented concurrent data structures that avoid this type
of locking.
However, since we normally use one queue per computational thread,
contention will only happens due to work-stealing, i.e.~when
another idle computational thread tries to poach tasks.
Since this happens only rarely, we opt for the simpler locking approach.
This decision is backed by the results in Section~5.

@@

The tasks are executed as follows:
\begin{center}\begin{minipage}{0.9\textwidth}
    \begin{lstlisting}
void qsched_run(qsched *s, void (*fun)(int, void *)) {
  qsched_start(s);
  #pragma omp parallel
  {
    int qid = omp_get_thread_num() % s->nr_queues;
    struct task *t;
    while ((t = qsched_gettask(s, qid)) != NULL) {
      fun(t->type, t->data);
      qsched_done(s, t);
    }
  }
}
    \end{lstlisting}
\end{minipage}\end{center}
\noindent where {\tt qsched\_start} initializes the tasks and
fills the queues (line~1).
For simplicity, OpenMP \citep{ref:Dagum1998}, which is available
for most compilers, is used to create a parallel section
in which the code between lines~4 and~11 is executed
concurrently.
A version using {\tt pthreads} \citep{ref:Pthreads1995}
directly\footnote{In most environments, OpenMP is implemented
on top of {\tt pthreads}, e.g. the {\tt gcc} compiler's libgomp.}
is also available.
The parallel section consists of a loop (lines~7--10) in
which a task is acquired via {\tt qsched\_gettask}
and its type and data are passed to a user-supplied
{\em execution function} {\tt fun}.
Once the task has been executed, it is returned to the
scheduler via the function {\tt qsched\_done}, i.e.~to
unlock its resources and unlock dependent tasks.
The loop terminates when the scheduler runs out of tasks,
i.e.~when {\tt qsched\_gettask} returns {\tt NULL}, and
the function exits once all the threads have exited their
loops.
